% =============================================================================
% NovaBASIC v1.0 — Copyright Page
% Emits content directly; \input{copyright} in the master document
% =============================================================================
\thispagestyle{empty}
\vspace*{\fill}
\begin{flushleft}

{\large\bfseries NovaBASIC v1.0 User Guide}\\[0.4em]
{\normalsize For the e6502 Virtual Computer --- 2026 Edition}

\bigskip
\noindent\rule{\textwidth}{0.4pt}
\bigskip

{\small
Copyright \textcopyright\ 2026 Barry Walker. All Rights Reserved.

\medskip
No part of this publication may be reproduced, distributed, or transmitted
in any form or by any means, including photocopying, recording, or other
electronic or mechanical methods, without the prior written permission of
the author, except in the case of brief quotations embodied in critical
reviews and certain other non-commercial uses permitted by copyright law.

\medskip
e6502 is an open source project.

\medskip
NovaBASIC is derived from Enhanced BASIC (EhBASIC) by Lee Davison.
EhBASIC is copyright \textcopyright\ Lee Davison and is used here under
the terms of its original free-for-non-commercial-use licence.
All extensions and modifications are the work of Barry Walker.

\medskip
MOS 6502 is a trademark of MOS Technology, Inc. All other trademarks
mentioned in this guide are the property of their respective owners.
Their use here is for identification purposes only.
}

\bigskip
\noindent\rule{\textwidth}{0.4pt}

\vspace{2em}
{\footnotesize\color{retrogray}
Typeset with \LaTeX\ and the \texttt{lmodern} font family.\\
Document class: \texttt{book}, 11\,pt, A4.\\
Build toolchain: \texttt{latexmk} with \texttt{pdflatex}.
}

\end{flushleft}
\vspace*{\fill}
