% =============================================================================
% Appendix A — VGC Bitmap & Sprite Shape RAM
% =============================================================================
\chapter{VGC Bitmap \& Sprite Shape RAM}

\epigraph{``The screen is a window into the machine's imagination.''}%
{\textit{--- Anonymous demo coder, 1989}}

\section*{Introduction}

The VGC manages two large memory spaces that live entirely outside the 6502's
64\,KB address space.  The graphics bitmap holds 64,000 bytes of pixel data for
the 320$\times$200 display, and the sprite shape RAM holds up to 256 shape
definitions of 128 bytes each (32\,KB total).  Neither space can be addressed
directly by the CPU; instead, the VGC provides two access mechanisms:

\begin{itemize}
  \item \textbf{MemRead/MemWrite commands} --- write a space selector and offset
        to the VGC command registers, then issue command \$19 (MemRead) or \$1A
        (MemWrite) at \addr{A010}.
  \item \textbf{DMA transfers} --- the DMA controller can copy blocks to and
        from VGC memory using space identifiers 3 (graphics bitmap) and 4
        (sprite shapes).
\end{itemize}


% ═══════════════════════════════════════════════════════════════════════════════
\section{Graphics Bitmap}
\label{sec:gfx-bitmap}
% ═══════════════════════════════════════════════════════════════════════════════

The graphics bitmap is a linear framebuffer of 320$\times$200 pixels stored in
row-major order.  Each pixel occupies one byte, with the colour index in the low
nibble (bits 3--0).  The upper nibble is ignored on write and reads as zero.

\begin{center}
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Property}         & \textbf{Value} \\
\midrule
Resolution                 & 320$\times$200 pixels \\
Bits per pixel             & 4 (stored in low nibble of each byte) \\
Total size                 & 64,000 bytes \\
VGC MemRead/MemWrite space & \$02 (\texttt{MemSpaceGfx}) \\
DMA space                  & 3 (\texttt{DmaSpaceVgcGfx}) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Address Calculation}

The offset of any pixel within the bitmap is:

\[
  \textit{offset} = y \times 320 + x
\]

where $x \in [0, 319]$ and $y \in [0, 199]$.

\subsection{Row Start Offsets}

The following table shows the starting offset (in decimal and hex) for every
10th row:

\begin{center}
\small
\begin{tabular}{@{}rrr@{\qquad}rrr@{}}
\toprule
\textbf{Row} & \textbf{Offset (dec)} & \textbf{Offset (hex)} &
\textbf{Row} & \textbf{Offset (dec)} & \textbf{Offset (hex)} \\
\midrule
  0 &     0 & \$0000 &  100 & 32000 & \$7D00 \\
 10 &  3200 & \$0C80 &  110 & 35200 & \$8980 \\
 20 &  6400 & \$1900 &  120 & 38400 & \$9600 \\
 30 &  9600 & \$2580 &  130 & 41600 & \$A280 \\
 40 & 12800 & \$3200 &  140 & 44800 & \$AF00 \\
 50 & 16000 & \$3E80 &  150 & 48000 & \$BB80 \\
 60 & 19200 & \$4B00 &  160 & 51200 & \$C800 \\
 70 & 22400 & \$5780 &  170 & 54400 & \$D480 \\
 80 & 25600 & \$6400 &  180 & 57600 & \$E100 \\
 90 & 28800 & \$7080 &  190 & 60800 & \$ED80 \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Drawing Commands vs.\ Direct Access}

Most programs use the VGC's built-in drawing commands (PLOT, LINE, CIRCLE, RECT,
FILL, PAINT) which operate on the bitmap internally.  The GCLS command clears
the entire bitmap to colour~0.  Direct MemRead/MemWrite access is useful for
reading individual pixel values or implementing custom rendering algorithms.

\subsection{BASIC Example: Read/Write Pixels via MemRead/MemWrite}

\begin{lstlisting}
10 REM read pixel at (100, 50)
20 POKE $A011, 2               : REM P0 = space (MemSpaceGfx)
30 OFFSET = 50*320+100
40 POKE $A012, OFFSET AND 255  : REM P1 = offset low byte
50 POKE $A013, INT(OFFSET/256) : REM P2 = offset high byte
60 POKE $A010, $19             : REM MemRead
70 PRINT "COLOR:"; PEEK($A014) : REM P3 = result
80 REM
90 REM write pixel at (100, 50) to color 1 (white)
100 POKE $A014, 1              : REM P3 = value to write
110 POKE $A010, $1A            : REM MemWrite
\end{lstlisting}

\begin{notebox}
The VGC command registers at \addr{A011}--\addr{A014} retain their values after
a command executes.  When reading or writing consecutive pixels, you only need to
update the offset registers between commands.
\end{notebox}


% ═══════════════════════════════════════════════════════════════════════════════
\section{Sprite Shape RAM}
\label{sec:sprite-shapes}
% ═══════════════════════════════════════════════════════════════════════════════

Sprite shape RAM holds up to 256 shape definitions.  Each shape is a
16$\times$16 pixel image stored as 128 bytes using 4-bit packed nibbles (two
pixels per byte).  The high nibble holds the even-column pixel and the low
nibble holds the odd-column pixel.  A pixel value of 0 is transparent by
default.

\begin{center}
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Property}         & \textbf{Value} \\
\midrule
Shape slots                & 256 (numbered 0--255) \\
Shape dimensions           & 16$\times$16 pixels \\
Bytes per row              & 8 (16 pixels $\times$ 4 bits $\div$ 8) \\
Bytes per shape            & 128 (8 bytes/row $\times$ 16 rows) \\
Bits per pixel             & 4 (packed: high nibble = even col, low nibble = odd col) \\
Total size                 & 32,768 bytes (32\,KB) \\
VGC MemRead/MemWrite space & \$03 (\texttt{MemSpaceSprite}) \\
DMA space                  & 4 (\texttt{DmaSpaceVgcSprite}) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Address Calculation}

The byte offset within sprite shape RAM is:

\[
  \textit{offset} = \textit{slot} \times 128 + \textit{row} \times 8 + \lfloor\textit{column} \div 2\rfloor
\]

where \textit{slot}~$\in [0, 255]$, \textit{row}~$\in [0, 15]$, and
\textit{column}~$\in [0, 15]$.  Within the byte, the even column occupies the
high nibble (bits~7--4) and the odd column occupies the low nibble (bits~3--0).

\subsection{Shape Slot Assignments}

\begin{notebox}
Shape slots 0--15 correspond to the default shapes for sprites 0--15 at
power-on, but any sprite can reference any shape slot by writing to the
\texttt{SprRegShape} register at offset~+4 within the sprite's register block
(base \addr{A040}).  This means multiple sprites can share a single shape
definition, and a sprite's appearance can be changed at any time by pointing it
to a different slot.
\end{notebox}

\subsection{BASIC Example: Define a Shape via MemWrite}

\begin{lstlisting}
10 REM define a simple arrow shape in slot 0
20 REM sprites are 4-bit packed: 8 bytes/row, 128 bytes/shape
30 POKE $A011, 3               : REM P0 = space (MemSpaceSprite)
40 FOR ROW = 0 TO 15
50   FOR BYTECOL = 0 TO 7
60     OFFSET = 0*128 + ROW*8 + BYTECOL
70     POKE $A012, OFFSET AND 255  : REM P1 = offset low
80     POKE $A013, INT(OFFSET/256) : REM P2 = offset high
90     HI = 0 : LO = 0 : REM two pixels per byte
100    LCOL = BYTECOL*2 : RCOL = LCOL+1
110    IF LCOL = 6 AND ROW < 12 THEN HI = 1
120    IF RCOL = 7 AND ROW < 12 THEN LO = 1
130    IF ROW = 0 AND LCOL >= 4 AND LCOL <= 8 THEN HI = 1
140    IF ROW = 0 AND RCOL >= 5 AND RCOL <= 9 THEN LO = 1
150    POKE $A014, HI*16 + LO     : REM P3 = packed nibbles
160    POKE $A010, $1A             : REM MemWrite
170  NEXT BYTECOL
180 NEXT ROW
\end{lstlisting}

\subsection{Using DMA for Bulk Shape Loading}

For loading many shapes at once (e.g., from XRAM or CPU RAM), the DMA controller
is far more efficient than individual MemWrite commands.  Set the DMA source
space to CPU RAM (space~0) or XRAM (space~5), the destination space to
\texttt{DmaSpaceVgcSprite} (space~4), and issue a DMA copy command.  See
Chapter~\ref{sec:xmc-registers} for DMA register details.

\begin{lstlisting}
10 REM DMA copy 1280 bytes (10 shapes x 128) from CPU RAM
20 REM at $2000 into sprite shape slot 0
30 DMACOPY $2000, 0, 1280, 0, 4
\end{lstlisting}
