% =============================================================================
% Chapter 8 — Screen Memory ($AA00–$B99F)
% =============================================================================
\chapter{Screen Memory}

\epigraph{``The screen is a window into the machine's mind.  What you write
there, the world will see.''}{\textit{--- An anonymous systems programmer}}

\section*{Introduction}

The NovaVM's 80$\times$25 text display is backed by two contiguous 2000-byte
RAM blocks: \textbf{Character RAM} holds the ASCII code for each cell, and
\textbf{Color RAM} holds the foreground colour.  Together they occupy
\$AA00--\$B99F --- 4000 bytes that map one-to-one to the 2000 visible character
positions on screen.

These addresses are ordinary read/write memory.  Any store to Character or
Color RAM takes effect immediately; there is no need to wait for VSYNC or issue
a VGC command.  This makes direct screen writes the fastest way to update small
regions of the display, ideal for scoreboards, status bars, and text
animations.


% ═══════════════════════════════════════════════════════════════════════════════
\section{Character RAM (\$AA00--\$B1CF)}
\label{sec:charram}
% ═══════════════════════════════════════════════════════════════════════════════

\memrange{AA00}{B1CF}{Character RAM}{R/W}{\$20}

2000 bytes arranged as 25 rows of 80 columns.  Each byte is a standard ASCII
character code; the VGC renders it using the built-in 8$\times$16 font.  The
default fill value is \$20 (space), giving a blank screen after reset or
\texttt{CLS}.

The address of any cell is calculated as:

\begin{center}
\texttt{addr = \$AA00 + row * 80 + column}
\end{center}

\noindent where \texttt{row} ranges from 0 (top) to 24 (bottom) and
\texttt{column} from 0 (left) to 79 (right).

Writing a byte to any address in this range instantly updates the
corresponding character on screen.  Reading returns the character code
currently displayed at that position.

\begin{lstlisting}
10 REM place an asterisk at column 40, row 12 (centre of screen)
20 POKE $AA00 + 12*80 + 40, ASC("*")
\end{lstlisting}


% ═══════════════════════════════════════════════════════════════════════════════
\section{Color RAM (\$B1D0--\$B99F)}
\label{sec:colorram}
% ═══════════════════════════════════════════════════════════════════════════════

\memrange{B1D0}{B99F}{Color RAM}{R/W}{1}

2000 bytes, one-to-one with Character RAM.  Each byte holds the foreground
colour (0--15) for the corresponding character cell.  The default value is~1
(white on the standard palette).

The address formula mirrors Character RAM:

\begin{center}
\texttt{addr = \$B1D0 + row * 80 + column}
\end{center}

Writing a colour value here changes the foreground colour of the
corresponding character immediately.  The background colour for all cells is
set globally via the VGC's \addr{A001} register (RegBgCol).

\begin{lstlisting}
10 REM set the character at (0,0) to red (colour 2)
20 POKE $B1D0, 2
\end{lstlisting}

\begin{notebox}
The VGC's 16-colour palette is documented in Appendix~D.  Colour index~0 is
black and index~1 is white by default, but the palette can be customised via
the VGC's colour registers.
\end{notebox}


% ═══════════════════════════════════════════════════════════════════════════════
\section{Screen Coordinate Math}
\label{sec:screen-math}
% ═══════════════════════════════════════════════════════════════════════════════

Each row occupies exactly 80 bytes (\$50 in hex).  The following table lists
the starting address of every row in both Character and Color RAM.

\begin{center}
\small
\begin{longtable}{@{}rll@{}}
\toprule
\textbf{Row} & \textbf{Char Addr} & \textbf{Color Addr} \\
\midrule
\endfirsthead
\toprule
\textbf{Row} & \textbf{Char Addr} & \textbf{Color Addr} \\
\midrule
\endhead
 0 & \texttt{\$AA00} & \texttt{\$B1D0} \\
 1 & \texttt{\$AA50} & \texttt{\$B220} \\
 2 & \texttt{\$AAA0} & \texttt{\$B270} \\
 3 & \texttt{\$AAF0} & \texttt{\$B2C0} \\
 4 & \texttt{\$AB40} & \texttt{\$B310} \\
 5 & \texttt{\$AB90} & \texttt{\$B360} \\
 6 & \texttt{\$ABE0} & \texttt{\$B3B0} \\
 7 & \texttt{\$AC30} & \texttt{\$B400} \\
 8 & \texttt{\$AC80} & \texttt{\$B450} \\
 9 & \texttt{\$ACD0} & \texttt{\$B4A0} \\
10 & \texttt{\$AD20} & \texttt{\$B4F0} \\
11 & \texttt{\$AD70} & \texttt{\$B540} \\
12 & \texttt{\$ADC0} & \texttt{\$B590} \\
13 & \texttt{\$AE10} & \texttt{\$B5E0} \\
14 & \texttt{\$AE60} & \texttt{\$B630} \\
15 & \texttt{\$AEB0} & \texttt{\$B680} \\
16 & \texttt{\$AF00} & \texttt{\$B6D0} \\
17 & \texttt{\$AF50} & \texttt{\$B720} \\
18 & \texttt{\$AFA0} & \texttt{\$B770} \\
19 & \texttt{\$AFF0} & \texttt{\$B7C0} \\
20 & \texttt{\$B040} & \texttt{\$B810} \\
21 & \texttt{\$B090} & \texttt{\$B860} \\
22 & \texttt{\$B0E0} & \texttt{\$B8B0} \\
23 & \texttt{\$B130} & \texttt{\$B900} \\
24 & \texttt{\$B180} & \texttt{\$B950} \\
\bottomrule
\end{longtable}
\end{center}

\subsection{Worked Example}

To place the letter \texttt{A} (ASCII \$41) in colour 14 (light blue) at
column~35, row~10:

\begin{enumerate}
  \item Character address: \$AA00 + 10 $\times$ 80 + 35 = \$AA00 + 835 =
        \$AA00 + \$0343 = \texttt{\$AD43}
  \item Color address: \$B1D0 + 10 $\times$ 80 + 35 = \$B1D0 + \$0343 =
        \texttt{\$B513}
\end{enumerate}

\begin{lstlisting}
POKE $AD43, $41 : POKE $B513, 14
\end{lstlisting}

\begin{tipbox}
For speed-critical routines, pre-compute row base addresses in an array rather
than multiplying by~80 on every access.  See the ``Fast Text Rendering''
example below.
\end{tipbox}


% ═══════════════════════════════════════════════════════════════════════════════
\section{BASIC Examples}
\label{sec:screen-examples}
% ═══════════════════════════════════════════════════════════════════════════════

\subsection{Direct Screen Write}

Write a string with a chosen colour without using \texttt{PRINT}:

\begin{lstlisting}
10 REM write "HELLO" at row 0, col 0 in red
20 H$ = "HELLO"
30 FOR I=0 TO LEN(H$)-1
40   POKE $AA00+I, ASC(MID$(H$,I+1,1))
50   POKE $B1D0+I, 2 : REM colour 2 = red
60 NEXT I
\end{lstlisting}

\subsection{Screen Dump}

Read the current contents of the screen and print them to a string:

\begin{lstlisting}
10 REM read row 0 of the screen
20 R$ = ""
30 FOR I=0 TO 79
40   C = PEEK($AA00+I)
50   IF C<32 THEN C=32
60   R$ = R$ + CHR$(C)
70 NEXT I
80 PRINT R$
\end{lstlisting}

\subsection{Fast Text Rendering}

For games and animations, direct POKEs outperform \texttt{PRINT} because they
avoid cursor movement, scrolling, and control-character interpretation.  This
example pre-computes row addresses for speed:

\begin{lstlisting}
100 REM pre-compute row base addresses
110 DIM CA(24), CO(24)
120 FOR R=0 TO 24
130   CA(R) = $AA00 + R*80
140   CO(R) = $B1D0 + R*80
150 NEXT R
200 REM draw a status bar on row 24
210 S$ = " SCORE: 00000   LIVES: 3   LEVEL: 01 "
220 FOR I=0 TO LEN(S$)-1
230   POKE CA(24)+I, ASC(MID$(S$,I+1,1))
240   POKE CO(24)+I, 14 : REM light blue
250 NEXT I
\end{lstlisting}

\subsection{Colour Cycling Effect}

Cycle all characters on screen through the 16 colours:

\begin{lstlisting}
10 REM colour cycling effect
20 C = 0
30 FOR I=0 TO 1999
40   POKE $B1D0+I, C
50 NEXT I
60 C = (C+1) AND 15
70 GOTO 30
\end{lstlisting}

\begin{notebox}
The VGC also provides indirect access to screen and color RAM via the
MemRead (\$19) and MemWrite (\$1A) commands (see \addr{A010}).  Direct
\texttt{POKE} is faster for small updates; DMA transfers are faster for bulk
operations.
\end{notebox}

\begin{warningbox}
Writing outside the 2000-byte range of each block will corrupt adjacent
hardware registers.  Character RAM ends at \$B1CF --- bytes written to
\$B1D0 and above will overwrite Color RAM.  Color RAM ends at \$B99F ---
bytes written to \$B9A0 and above will hit the File I/O Controller registers.
Always bounds-check your row and column values.
\end{warningbox}
